---
title: "Data Structures"
teaching: 40
exercises: 15
questions:
- "How can I read data in R?"
- "What are the basic data types in R?"
- "How do I represent categorical information in R?"
objectives:
- "To be aware of the different types of data."
- "To begin exploring data frames, and understand how they are related to vectors, factors and lists."
- "To be able to ask questions from R about the type, class, and structure of an object."
keypoints:
- "Use `read.csv` to read tabular data in R."
- "The basic data types in R are double, integer, complex, logical, and character."
- "Use factors to represent categories in R."
source: Rmd
---
## Using R as a calculator

The simplest thing you could do with R is do arithmetic:

```{r}
1 + 100
```

And R will print out the answer, with a preceding "[1]". Don't worry about this
for now, we'll explain that later. For now think of it as indicating output.

Like bash, if you type in an incomplete command, R will wait for you to
complete it:

~~~
> 1 +
~~~
{: .r}

~~~
+
~~~
{: .output}

Any time you hit return and the R session shows a "+" instead of a ">", it
means it's waiting for you to complete the command. If you want to cancel
a command you can simply hit "Esc" and RStudio will give you back the ">"
prompt.

> ## Tip: Cancelling commands
>
> If you're using R from the commandline instead of from within RStudio,
> you need to use <kbd>Ctrl</kbd>+<kbd>C</kbd> instead of <kbd>Esc</kbd>
> to cancel the command. This applies to Mac users as well!
>
> Cancelling a command isn't only useful for killing incomplete commands:
> you can also use it to tell R to stop running code (for example if it's
> taking much longer than you expect), or to get rid of the code you're
> currently writing.
>
{: .callout}

When using R as a calculator, the order of operations is the same as you
would have learned back in school.

From highest to lowest precedence:

 * Parentheses: `(`, `)`
 * Exponents: `^` or `**`
 * Divide: `/`
 * Multiply: `*`
 * Add: `+`
 * Subtract: `-`

```{r}
3 + 5 * 2
```

Use parentheses to group operations in order to force the order of
evaluation if it differs from the default, or to make clear what you
intend.

```{r}
(3 + 5) * 2
```

This can get unwieldy when not needed, but  clarifies your intentions.
Remember that others may later read your code.

```{r, eval=FALSE}
(3 + (5 * (2 ^ 2))) # hard to read
3 + 5 * 2 ^ 2       # clear, if you remember the rules
3 + 5 * (2 ^ 2)     # if you forget some rules, this might help
```


The text after each line of code is called a
"comment". Anything that follows after the hash (or octothorpe) symbol
`#` is ignored by R when it executes code.

Really small or large numbers get a scientific notation:

```{r}
2/10000
```

Which is shorthand for "multiplied by `10^XX`". So `2e-4`
is shorthand for `2 * 10^(-4)`.

You can write numbers in scientific notation too:

```{r}
5e3  # Note the lack of minus here
```

## Mathematical functions

R has many built in mathematical functions. To call a function,
we simply type its name, followed by  open and closing parentheses.
Anything we type inside the parentheses is called the function's
arguments:

```{r}
sin(1)  # trigonometry functions
```

```{r}
log(1)  # natural logarithm
```

```{r}
log10(10) # base-10 logarithm
```

```{r}
exp(0.5) # e^(1/2)
```

Don't worry about trying to remember every function in R. You
can simply look them up on Google, or if you can remember the
start of the function's name, use the tab completion in RStudio.

This is one advantage that RStudio has over R on its own, it
has auto-completion abilities that allow you to more easily
look up functions, their arguments, and the values that they
take.

Typing a `?` before the name of a command will open the help page
for that command. As well as providing a detailed description of
the command and how it works, scrolling to the bottom of the
help page will usually show a collection of code examples which
illustrate command usage. We'll go through an example later.

## Comparing things

We can also do comparison in R:

```{r}
1 == 1  # equality (note two equals signs, read as "is equal to")
```

```{r}
1 != 2  # inequality (read as "is not equal to")
```

```{r}
1 < 2  # less than
```

```{r}
1 <= 1  # less than or equal to
```

```{r}
1 > 0  # greater than
```

```{r}
1 >= -9 # greater than or equal to
```

> ## Tip: Comparing Numbers
>
> A word of warning about comparing numbers: you should
> never use `==` to compare two numbers unless they are
> integers (a data type which can specifically represent
> only whole numbers).
>
> Computers may only represent decimal numbers with a
> certain degree of precision, so two numbers which look
> the same when printed out by R, may actually have
> different underlying representations and therefore be
> different by a small margin of error (called Machine
> numeric tolerance).
>
> Instead you should use the `all.equal` function.
>
> Further reading: [http://floating-point-gui.de/](http://floating-point-gui.de/)
>
{: .callout}

## Variables and assignment

We can store values in variables using the assignment operator `<-`, like this:

```{r}
x <- 1/40
```

Notice that assignment does not print a value. Instead, we stored it for later
in something called a **variable**. `x` now contains the **value** `0.025`:

```{r}
x
```

More precisely, the stored value is a *decimal approximation* of
this fraction called a [floating point number](http://en.wikipedia.org/wiki/Floating_point).

Look for the `Environment` tab in one of the panes of RStudio, and you will see that `x` and its value
have appeared. Our variable `x` can be used in place of a number in any calculation that expects a number:

```{r}
log(x)
```

Notice also that variables can be reassigned:

```{r}
x <- 100
```

`x` used to contain the value 0.025 and and now it has the value 100.

Assignment values can contain the variable being assigned to:

```{r}
x <- x + 1 #notice how RStudio updates its description of x on the top right tab
y <- x * 2
```

The right hand side of the assignment can be any valid R expression.
The right hand side is *fully evaluated* before the assignment occurs.

Variable names can contain letters, numbers, underscores and periods. They
cannot start with a number nor contain spaces at all. Different people use
different conventions for long variable names, these include

  * periods.between.words
  * underscores\_between_words
  * camelCaseToSeparateWords

What you use is up to you, but **be consistent**.

> ## Challenge 1
>
> Which of the following are valid R variable names?
> ```{r, eval=FALSE}
> min_height
> max.height
> _age
> .mass
> MaxLength
> min-length
> 2widths
> celsius2kelvin
> ```
>
> > ## Solution to challenge 1
> >
> > The following can be used as R variables:
> > ```{r ch1pt1-sol, eval=FALSE}
> > min_height
> > max.height
> > MaxLength
> > celsius2kelvin
> > ```
> >
> > The following creates a hidden variable:
> > ```{r ch1pt2-sol, eval=FALSE}
> > .mass
> > ```
> >
> > The following will not be able to be used to create a variable
> > ```{r ch1pt3-sol, eval=FALSE}
> > _age
> > min-length
> > 2widths
> > ```
> {: .solution}
{: .challenge}

## Vectorization

One final thing to be aware of is that R is *vectorized*, meaning that
variables and functions can have vectors as values. In contrast to physics and
mathematics, a vector in R describes a set of values in a certain order of the 
same data type. For example

```{r}
1:5
2^(1:5)
x <- 1:5
2^x
```

This is incredibly powerful; we will discuss this further in an
upcoming lesson.


## Managing your environment

There are a few useful commands you can use to interact with the R session.

`ls` will list all of the variables and functions stored in the global environment
(your working R session):

```{r}
ls()
```

> ## Tip: hidden objects
>
> Like in the shell, `ls` will hide any variables or functions starting
> with a "." by default. To list all objects, type `ls(all.names=TRUE)`
> instead
>
{: .callout}

Note here that we didn't give any arguments to `ls`, but we still
needed to give the parentheses to tell R to call the function.

If we type `ls` by itself, R will print out the source code for that function!

```{r}
ls
```

You can use `rm` to delete objects you no longer need:

```{r, eval=FALSE}
rm(x)
```

If you have lots of things in your environment and want to delete all of them,
you can pass the results of `ls` to the `rm` function:

```{r, eval=FALSE}
rm(list = ls())
```

In this case we've combined the two. Like the order of operations, anything
inside the innermost parentheses is evaluated first, and so on.

In this case we've specified that the results of `ls` should be used for the
`list` argument in `rm`. When assigning values to arguments by name, you *must*
use the `=` operator!!

If instead we use `<-`, there will be unintended side effects, or you may get an error message:

```{r, error=TRUE}
rm(list <- ls())
```

> ## Tip: Warnings vs. Errors
>
> Pay attention when R does something unexpected! Errors, like above,
> are thrown when R cannot proceed with a calculation. Warnings on the
> other hand usually mean that the function has run, but it probably
> hasn't worked as expected.
>
> In both cases, the message that R prints out usually give you clues
> how to fix a problem.
>
{: .callout}


## data types 

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("04-")

## Save the data
cats_orig <- data.frame(coat = factor(c("calico", "black", "tabby")),
                   weight = c(2.1, 5.0, 3.2),
                   likes_string = c(1, 0, 1))
cats_bad <- data.frame(coat = factor(c("calico", "black", "tabby", "tabby")),
                   weight = factor(c(2.1, 5.0, 3.2, '2.3 or 2.4')),
                   likes_string = c(1, 0, 1, 1))
cats <- cats_orig
```

One of R's most powerful features is its ability to deal with tabular data -
such as you may already have in a spreadsheet or a CSV file. Let's start by
making a toy dataset in your `data/` directory, called `feline-data.csv`:

```{r}
cats <- data.frame(coat = c("calico", "black", "tabby"), 
                    weight = c(2.1, 5.0,3.2), 
                    likes_string = c(1, 0, 1))
write.csv(x = cats, file = "data/feline-data.csv", row.names = FALSE)
```
The contents of the new file, `feline-data.csv`:
```{r, eval=FALSE}
coat,weight,likes_string
calico,2.1,1
black,5.0,0
tabby,3.2,1
```

> ## Tip: Editing Text files in R
>
> Alternatively, you can create `data/feline-data.csv` using a text editor (Nano),
> or within RStudio with the **File -> New File -> Text File** menu item.
{: .callout}

We can load this into R via the following:

```{r}
cats <- read.csv(file = "data/feline-data.csv")
cats
```

The `read.table` function is used for reading in tabular data stored in a text
file where the columns of data are separated by punctuation characters such as
CSV files (csv = comma-separated values). Tabs and commas are the most common
punctuation characters used to separate or delimit data points in csv files. 
For convenience R provides 2 other versions of `read.table`. These are: `read.csv`
for files where the data are separated with commas and `read.delim` for files
where the data are separated with tabs. Of these three functions `read.csv` is
the most commonly used.  If needed it is possible to override the default 
delimiting punctuation marks for both `read.csv` and `read.delim`.


We can begin exploring our dataset right away, pulling out columns by specifying
them using the `$` operator:

```{r}
cats$weight
cats$coat
```

We can do other operations on the columns:

```{r}
## Say we discovered that the scale weighs two Kg light:
cats$weight + 2
paste("My cat is", cats$coat)
```

But what about

```{r}
cats$weight + cats$coat
```

Understanding what happened here is key to successfully analyzing data in R.

## Data Types

If you guessed that the last command will return an error because `2.1` plus
`"black"` is nonsense, you're right - and you already have some intuition for an
important concept in programming called *data types*. We can ask what type of
data something is:

```{r}
typeof(cats$weight)
```

There are 5 main types: `double`, `integer`, `complex`, `logical` and `character`.

```{r}
typeof(3.14)
typeof(1L) # The L suffix forces the number to be an integer, since by default R uses float numbers
typeof(1+1i)
typeof(TRUE)
typeof('banana')
```

No matter how
complicated our analyses become, all data in R is interpreted as one of these
basic data types. This strictness has some really important consequences.

A user has added details of another cat. This information is in the file
`data/feline-data_v2.csv`.


```{r, eval=FALSE}
file.show("data/feline-data_v2.csv")
```

```{r, eval=FALSE}
coat,weight,likes_string
calico,2.1,1
black,5.0,0
tabby,3.2,1
tabby,2.3 or 2.4,1
```

Load the new cats data like before, and check what type of data we find in the
`weight` column:

```{r}
cats <- read.csv(file="data/feline-data_v2.csv")
typeof(cats$weight)
```

Oh no, our weights aren't the double type anymore! If we try to do the same math
we did on them before, we run into trouble:

```{r}
cats$weight + 2
```

What happened? When R reads a csv file into one of these tables, it insists that
everything in a column be the same basic type; if it can't understand
*everything* in the column as a double, then *nobody* in the column gets to be a
double. The table that R loaded our cats data into is something called a
*data.frame*, and it is our first example of something called a *data
structure* - that is, a structure which R knows how to build out of the basic
data types.

We can see that it is a *data.frame* by calling the `class` function on it:

```{r}
class(cats)
```

In order to successfully use our data in R, we need to understand what the basic
data structures are, and how they behave. For now, let's remove that extra line
from our cats data and reload it, while we investigate this behavior further:

feline-data.csv:

```
coat,weight,likes_string
calico,2.1,1
black,5.0,0
tabby,3.2,1
```

And back in RStudio:

```{r, eval=FALSE}
cats <- read.csv(file="data/feline-data.csv")
```

```{r, include=FALSE}
cats <- cats_orig
```


## Vectors and Type Coercion

To better understand this behavior, let's meet another of the data structures:
the *vector*.

```{r}
my_vector <- vector(length = 3)
my_vector
```

A vector in R is essentially an ordered list of things, with the special
condition that *everything in the vector must be the same basic data type*. If
you don't choose the datatype, it'll default to `logical`; or, you can declare
an empty vector of whatever type you like.


```{r}
another_vector <- vector(mode='character', length=3)
another_vector
```

You can check if something is a vector:

```{r}
str(another_vector)
```

The somewhat cryptic output from this command indicates the basic data type
found in this vector - in this case `chr`, character; an indication of the
number of things in the vector - actually, the indexes of the vector, in this
case `[1:3]`; and a few examples of what's actually in the vector - in this case
empty character strings. If we similarly do

```{r}
str(cats$weight)
```

we see that `cats$weight` is a vector, too - *the columns of data we load into R
data.frames are all vectors*, and that's the root of why R forces everything in
a column to be the same basic data type.

> ## Discussion 1
>
> Why is R so opinionated about what we put in our columns of data?
> How does this help us?
>
> > ## Discussion 1
> >
> > By keeping everything in a column the same, we allow ourselves to make simple
> > assumptions about our data; if you can interpret one entry in the column as a
> > number, then you can interpret *all* of them as numbers, so we don't have to
> > check every time. This consistency is what people mean when they talk about 
> > *clean data*; in the long run, strict consistency goes a long way to making 
> > our lives easier in R.
> {: .solution}
{: .discussion}

You can also make vectors with explicit contents with the combine function:

```{r}
combine_vector <- c(2,6,3)
combine_vector
```

Given what we've learned so far, what do you think the following will produce?

```{r}
quiz_vector <- c(2,6,'3')
```

This is something called *type coercion*, and it is the source of many surprises
and the reason why we need to be aware of the basic data types and how R will
interpret them. When R encounters a mix of types (here numeric and character) to
be combined into a single vector, it will force them all to be the same
type. Consider:

```{r}
coercion_vector <- c('a', TRUE)
coercion_vector
another_coercion_vector <- c(0, TRUE)
another_coercion_vector
```

The coercion rules go: `logical` -> `integer` -> `numeric` -> `complex` ->
`character`, where -> can be read as *are transformed into*. You can try to
force coercion against this flow using the `as.` functions:

```{r}
character_vector_example <- c('0','2','4')
character_vector_example
character_coerced_to_numeric <- as.numeric(character_vector_example)
character_coerced_to_numeric
numeric_coerced_to_logical <- as.logical(character_coerced_to_numeric)
numeric_coerced_to_logical
```

As you can see, some surprising things can happen when R forces one basic data
type into another! Nitty-gritty of type coercion aside, the point is: if your
data doesn't look like what you thought it was going to look like, type coercion
may well be to blame; make sure everything is the same type in your vectors and
your columns of data.frames, or you will get nasty surprises!

But coercion can also be very useful! For example, in our `cats` data
`likes_string` is numeric, but we know that the 1s and 0s actually represent
`TRUE` and `FALSE` (a common way of representing them). We should use the
`logical` datatype here, which has two states: `TRUE` or `FALSE`, which is
exactly what our data represents. We can 'coerce' this column to be `logical` by
using the `as.logical` function:

```{r}
cats$likes_string
cats$likes_string <- as.logical(cats$likes_string)
cats$likes_string
```

The combine function, `c()`, will also append things to an existing vector:

```{r}
ab_vector <- c('a', 'b')
ab_vector
combine_example <- c(ab_vector, 'SWC')
combine_example
```

You can also make series of numbers:

```{r}
mySeries <- 1:10
mySeries
seq(10)
seq(1,10, by=0.1)
```

We can ask a few questions about vectors:

```{r}
sequence_example <- seq(10)
head(sequence_example, n=2)
tail(sequence_example, n=4)
length(sequence_example)
class(sequence_example)
typeof(sequence_example)
```

Finally, you can give names to elements in your vector:

```{r}
my_example <- 5:8
names(my_example) <- c("a", "b", "c", "d")
my_example
names(my_example)
```

> ## Challenge 1
>
> Start by making a vector with the numbers 1 through 26.
> Multiply the vector by 2, and give the resulting vector
> names A through Z (hint: there is a built in vector called `LETTERS`)
>
> > ## Solution to Challenge 1
> >
> > ```{r}
> > x <- 1:26
> > x <- x * 2
> > names(x) <- LETTERS
> > ```
> {: .solution}
{: .challenge}


## Data Frames

We said that columns in data.frames were vectors:

```{r}
str(cats$weight)
str(cats$likes_string)
```

These make sense. But what about

```{r}
str(cats$coat)
```

## Factors

Another important data structure is called a *factor*. Factors usually look like
character data, but are typically used to represent categorical information. For
example, let's make a vector of strings labelling cat colorations for all the
cats in our study:

```{r}
coats <- c('tabby', 'tortoiseshell', 'tortoiseshell', 'black', 'tabby')
coats
str(coats)
```

We can turn a vector into a factor like so:

```{r}
CATegories <- factor(coats)
class(CATegories)
str(CATegories)
```

Now R has noticed that there are three possible categories in our data - but it
also did something surprising; instead of printing out the strings we gave it,
we got a bunch of numbers instead. R has replaced our human-readable categories
with numbered indices under the hood, this is necessary as many statistical
calculations utilise such numerical representations for categorical data:

```{r}
typeof(coats)
typeof(CATegories)
```

> ## Challenge 2
>
> Is there a factor in our `cats` data.frame? what is its name?
> Try using `?read.csv` to figure out how to keep text columns as character
> vectors instead of factors; then write a command or two to show that the factor
> in `cats` is actually a character vector when loaded in this way.
>
> > ## Solution to Challenge 2
> >
> > One solution is use the argument `stringAsFactors`:
> >
> > ```{r, eval = FALSE}
> > cats <- read.csv(file="data/feline-data.csv", stringsAsFactors=FALSE)
> > str(cats$coat)
> > ```
> >
> > Another solution is use the argument `colClasses`
> > that allow finer control.
> >
> > ```{r, eval = FALSE}
> > cats <- read.csv(file="data/feline-data.csv", colClasses=c(NA, NA, "character"))
> > str(cats$coat)
> > ```
> >
> > Note: new students find the help files difficult to understand; make sure to let them know
> > that this is typical, and encourage them to take their best guess based on semantic meaning,
> > even if they aren't sure.
> {: .solution}
{: .challenge}

In modelling functions, it's important to know what the baseline levels are. This
is assumed to be the first factor, but by default factors are labelled in
alphabetical order. You can change this by specifying the levels:

```{r}
mydata <- c("case", "control", "control", "case")
factor_ordering_example <- factor(mydata, levels = c("control", "case"))
str(factor_ordering_example)
```

In this case, we've explicitly told R that "control" should represented by 1, and
"case" by 2. This designation can be very important for interpreting the
results of statistical models!

## Lists

Another data structure you'll want in your bag of tricks is the `list`. A list
is simpler in some ways than the other types, because you can put anything you
want in it:

```{r}
list_example <- list(1, "a", TRUE, 1+4i)
list_example
another_list <- list(title = "Numbers", numbers = 1:10, data = TRUE )
another_list
```

We can now understand something a bit surprising in our data.frame; what happens if we run:

```{r}
typeof(cats)
```

We see that data.frames look like lists 'under the hood' - this is because a
data.frame is really a list of vectors and factors, as they have to be - in
order to hold those columns that are a mix of vectors and factors, the
data.frame needs something a bit more flexible than a vector to put all the
columns together into a familiar table.  In other words, a `data.frame` is a
special list in which all the vectors must have the same length.

In our `cats` example, we have an integer, a double and a logical variable. As
we have seen already, each column of data.frame is a vector.

```{r}
cats$coat
cats[,1]
typeof(cats[,1])
str(cats[,1])
```

Each row is an *observation* of different variables, itself a data.frame, and
thus can be composed of elements of different types.

```{r}
cats[1,]
typeof(cats[1,])
str(cats[1,])
```

> ## Challenge 3
>
> There are several subtly different ways to call variables, observations and
> elements from data.frames:
>
> - `cats[1]`
> - `cats[[1]]`
> - `cats$coat`
> - `cats["coat"]`
> - `cats[1, 1]`
> - `cats[, 1]`
> - `cats[1, ]`
>
> Try out these examples and explain what is returned by each one.
>
> *Hint:* Use the function `typeof()` to examine what is returned in each case.
>
> > ## Solution to Challenge 3
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[1]
> > ```
> > We can think of a data frame as a list of vectors. The single brace `[1]`
> returns the first slice of the list, as another list. In this case it is the
> first column of the data frame.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[[1]]
> > ```
> > The double brace `[[1]]` returns the contents of the list item. In this case
> it is the contents of the first column, a _vector_ of type _factor_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats$coat
> > ```
> > This example uses the `$` character to address items by name. _coat_ is the
> first column of the data frame, again a _vector_ of type _factor_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats["coat"]
> > ```
> > Here we are using a single brace `["coat"]` replacing the index number with
> the column name. Like example 1, the returned object is a _list_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[1, 1]
> > ```
> > This example uses a single brace, but this time we provide row and column
> coordinates. The returned object is the value in row 1, column 1. The object
> is an _integer_ but because it is part of a _vector_ of type _factor_, R
> displays the label "calico" associated with the integer value.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[, 1]
> > ```
> > Like the previous example we use single braces and provide row and column
> coordinates. The row coordinate is not specified, R interprets this missing
> value as all the elements in this _column_ _vector_.
> > ```{r, eval=TRUE, echo=TRUE}
> > cats[1, ]
> > ```
> > Again we use the single brace with row and column coordinates. The column
> coordinate is not specified. The return value is a _list_ containing all the
> values in the first row.
> {: .solution}
{: .challenge}

## Matrices

Last but not least is the matrix. We can declare a matrix full of zeros:

```{r}
matrix_example <- matrix(0, ncol=6, nrow=3)
matrix_example
```

And similar to other data structures, we can ask things about our matrix:

```{r}
class(matrix_example)
typeof(matrix_example)
str(matrix_example)
dim(matrix_example)
nrow(matrix_example)
ncol(matrix_example)
```