---
title: "Data Structures"
teaching: 20
exercises: 15
questions:
- "How can I read data in R?"
- "How can I create a variable in R?"
objectives:
- "To be able to ask questions from R about the type, class, and structure of an object."
keypoints:
- "Use `read.csv` to read tabular data in R."
- "The basic data types in R are double, integer, complex, logical, and character."
- "Use factors to represent categories in R."
source: Rmd
---
## Using R as a calculator

The simplest thing you could do with R is do arithmetic:

```{r}
1 + 100
```

And R will print out the answer, with a preceding "[1]". Don't worry about this
for now, we'll explain that later. For now think of it as indicating output.

Like bash, if you type in an incomplete command, R will wait for you to
complete it:

~~~
> 1 +
~~~
{: .r}

~~~
+
~~~
{: .output}

Any time you hit return and the R session shows a "+" instead of a ">", it
means it's waiting for you to complete the command. If you want to cancel
a command you can simply hit "Esc" and RStudio will give you back the ">"
prompt.

> ## Tip: Cancelling commands
>
> If you're using R from the commandline instead of from within RStudio,
> you need to use <kbd>Ctrl</kbd>+<kbd>C</kbd> instead of <kbd>Esc</kbd>
> to cancel the command. This applies to Mac users as well!
>
> Cancelling a command isn't only useful for killing incomplete commands:
> you can also use it to tell R to stop running code (for example if it's
> taking much longer than you expect), or to get rid of the code you're
> currently writing.
>
{: .callout}

When using R as a calculator, the order of operations is the same as you
would have learned back in school.

From highest to lowest precedence:

 * Parentheses: `(`, `)`
 * Exponents: `^` or `**`
 * Divide: `/`
 * Multiply: `*`
 * Add: `+`
 * Subtract: `-`

```{r}
3 + 5 * 2
```

Use parentheses to group operations in order to force the order of
evaluation if it differs from the default, or to make clear what you
intend.

```{r}
(3 + 5) * 2
```

Really small or large numbers get a scientific notation:

```{r}
2/10000
```

Which is shorthand for "multiplied by `10^XX`". So `2e-4`
is shorthand for `2 * 10^(-4)`.

You can write numbers in scientific notation too:

```{r}
5e3  # Note the lack of minus here
```

## Mathematical functions

R has many built in mathematical functions. To call a function,
we simply type its name, followed by  open and closing parentheses.
Anything we type inside the parentheses is called the function's
arguments:

```{r}
sin(1)  # trigonometry functions
```

```{r}
log(1)  # natural logarithm
```


* Don't worry about trying to remember every function in R. 
* Look them up on Google, or if you can remember the
start of the function's name, use the tab completion in RStudio.

### Get help 

* Typing a `?` before the name of a command will open the help page
for that command. 
* As well as providing a detailed description of
the command and how it works, scrolling to the bottom of the
help page will usually show a collection of code examples which
illustrate command usage.

## Comparing things

We can also do comparison in R:

```{r}
1 == 1  # equality (note two equals signs, read as "is equal to")
```

```{r}
1 != 2  # inequality (read as "is not equal to")
```

```{r}
1 < 2  # less than
```

```{r}
1 <= 1  # less than or equal to
```


## Variables and assignment

* We can store values in variables using the assignment operator `<-`, like this. 
* Short cut is Windows/Linux: <kAlt" + "-"
Mac: "Option" + "-"

```{r}
x <- 1/40
```

Notice that assignment does not print a value. Instead, we stored it for later
in something called a **variable**. `x` now contains the **value** `0.025`:

```{r}
x
```

Notice also that variables can be reassigned:

```{r}
x <- 100
```

`x` used to contain the value 0.025 and and now it has the value 100.

Assignment values can contain the variable being assigned to:

```{r}
x <- x + 1 #notice how RStudio updates its description of x on the top right tab
y <- x * 2
```

The right hand side of the assignment can be any valid R expression.
The right hand side is *fully evaluated* before the assignment occurs.

Variable names can contain letters, numbers, underscores and periods. They
cannot start with a number nor contain spaces at all. Different people use
different conventions for long variable names, these include

  * periods.between.words
  * underscores\_between_words
  * camelCaseToSeparateWords

What you use is up to you, but **be consistent**.

> ## Challenge 1
>
> Which of the following are valid R variable names?
> ```{r, eval=FALSE}
> min_height
> max.height
> _age
> .mass
> MaxLength
> min-length
> 2widths
> celsius2kelvin
> ```
>
> > ## Solution to challenge 1
> >
> > The following can be used as R variables:
> > ```{r ch1pt1-sol, eval=FALSE}
> > min_height
> > max.height
> > MaxLength
> > celsius2kelvin
> > ```
> >
> > The following creates a hidden variable:
> > ```{r ch1pt2-sol, eval=FALSE}
> > .mass
> > ```
> >
> > The following will not be able to be used to create a variable
> > ```{r ch1pt3-sol, eval=FALSE}
> > _age
> > min-length
> > 2widths
> > ```
> {: .solution}
{: .challenge}

## Getting Data into R

