---
title: "Reliability Data Preparation"
author: "Tim Dennis"
date: "2019-01-27"
output: html_document
teaching: 20
exercises: 
questions:
- "How do I prep data for analysis?"
- "How can I merge data from one dataset to another?"
objectives:
- "To be able to use `dplyr` to filter our dataframe by characeristics in it."
- "To be able to create a computed variable based on existing variables."
- "To use a changing structure to prepare our data for analysis."
- "To learn how to join data from another table into our analysis data."
- "To save a plot created with ggplot to disk."
keypoints:
- "Use `dplyr` to clean up our dataset."
- "Use `dplyr` to filter our data"
source: Rmd
---

```{r setup, include=FALSE}
# Install ggplot and ggmap (this is only required once)
#install.packages('ggplot2', dependencies = TRUE)
#install.packages('ggmap', dependencies = TRUE)
```

```{r packages}
# load packages
library(ggplot2)
library(ggmap)
library(readr)
library(dplyr)
library(stringr)
```

## R Markdown

We are now going to prepare our reliabitliy data for analysis and mapping by filtering it by type of service, create a reliability metric based on a calculations of two variables in our datset, and then pulling in latitude and longitude from another table. 

To begin let's read in the datasets we will need to work with. This will be two:

1. `TDashboardData_reliability_20160301-20160331.csv` downloaded via mbta online
2. `mbta_stations.csv` provided by ...




```{r}
#read in MBTA performance csv file
rawdata <- read_csv("data/TDashboardData_reliability_20160301-20160331.csv")
stations <- read_csv("data/mbta_stations.csv")
View(rawdata)
View(stations)
```

We first want to work with the reliability data.

1. Take rawdata and pipe through filter
1. Filter by peak service (weekday rush hour)
1. Filter by only the green line
1. Mutate to create or overwrite a variable 
1. str_trim to remove whitespaces before and after values
1. Create a rely variable that 
1. Finds the reliability at each station
1. Divides the numerator (people who have to wait too long) by the denominator (all riders).
1. This is the percentage of people who had to wait. 1 - this ratio is the percentage of people who didn't.
1. drop the columns we don't need for futher analysis

```{r}
rawdata %>% 
  filter(PEAK_OFFPEAK_IND == "Peak Service (Weekdays 6:30-9:30AM, 3:30PM-6:30PM)", ROUTE_TYPE=="Green Line" ) %>% 
  mutate(STOP = str_trim(STOP), rely = 1 - OTP_NUMERATOR/OTP_DENOMINATOR, rely = round(rely, 4)) %>% 
  select(-c("SERVICE_DATE","ROUTE_TYPE")) %>% 
  
```

## Adding Geolocation data to the dataframe via another table

* The lat/long data is in mbta_stations.csv 
* Import the MBTA station location data
* We will set up our dplyr sequence taking `stations` and passing it thru funcitons
* Select only the Green Line data


locs <- stations[which (stations$LINE=='GREEN'),]

# Change the name of Station to Stop to match the other dataframe
colnames(locs)[2] <- "STOP" 
View(locs)


# Join the location data to the reliability file
joindata <- merge(x = reliability, y = locs, by ="STOP", all.x=TRUE)

#Select only the columns need
joindata <- joindata[c("STOP","SERVICE_DATE","ROUTE_TYPE","rely","LONGITUDE","LATITUDE")]
x = !duplicated(joindata$STOP)
joindata <- joindata[x ,]
View(joindata)

