---
title: "Append csv Files into One Dataframe"
author: "Tim Dennis"
date: "2019-01-27"
output: html_document
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("09-")
```

## Load needed package

```{r load-packages}
library(tidyverse)
```

* The problem is that we have a number of ridership files by date as csv. We want to get those together into one. In this example we'll use the `purrr` package `map_dfr` function to return data frames created by **row-binding** together the files. Note that this is possible b/c the csv files are the same shape: have identical columns, etc.

```{r}
# List all files ending with csv in directory
csv_files = list.files(path = 'data/mbtaridership_2016/', pattern = "csv$", full.names = TRUE)
# Read each csv file into a list
ridership <- map_dfr(csv_files, read_csv)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(ridership)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
